Causal Discovery on Precarity and Depression

### 2025-01-04 draft_v1/draft_v1.qmd
update_discussion_limitations

# Results
## Depression as sum score
The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precarity factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*) under two different setups: (a) using both Gaussian CI test and RCoT, and (b) using RCoT alone. Black edges represent consistent causal relationships identified by both FCI and CCI, while gray edges denote inconsistent relationships that vary between the two methods. The edge endpoints of inconsistent gray edges are marked with circles.
In graph (a), the key pathways suggest that employment precarity (*P.emp*) and social precarity (*P.soc*) do not cause depression (*PHQsum*). Social precarity does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder. Additionally, *P.emp* and *S.fin* are identified as non-causes of housing precarity (*P.hou*). Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or if the connection is mediated by an unmeasured confounder. A similar ambiguity exists in the relationships between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*, with the methods diverging on whether these stressors influence depression or are linked through latent variables.

Graph (b), derived solely from the nonparametric RCoT test. Both graphs consistently identify that employment precarity is not a cause of housing precarity or depression, and that social precarity does not cause depression or financial stress. 
The relationship between stressors and depression remains unresolved between the two algorithms. However, RCoT provides greater confidence that depression is not the cause of financial stress and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of recent stressors as being closely linked with depression, either as causes or through latent confounders. Employment and social precarity may also be influenced by depression, either directly or through latent variables. Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, either directly or through a latent confounder.

@@

The present study explored the causal relationships between precarity factors and depression by employing both sum score and symptom-level analyses using causal discovery algorithms. The use of both PHQ sum scores and individual symptom scores provided a comprehensive understanding of how different aspects of precarity influence depression, revealing nuanced pathways that may be obscured when considering aggregate measures alone.

@@

<!-- Limitations -->
Several limitations should be acknowledged when interpreting these findings. The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data. These uncertainties underscore the need for further research, ideally incorporating datasets with a higher proportion of symptomatic individuals. The HELIUS dataset, being predominantly composed of asymptomatic samples, posed challenges in identifying clear causal directions, particularly among symptom nodes.
Future studies could address these limitations by incorporating time-series data to leverage temporal information about the relationships between precarity factors and depressive symptoms. Time-series data could provide time-specific insights and track how these relationships evolve over time. For instance, methods such as *PCMCI* [@runge2019detecting] and *tsFCI* [@entner2010causal], along with other time-series adaptations of causal discovery algorithms, could help better account for temporal dependencies and refine the analysis.

Another limitation is the lack of clear evidence for cycles within the symptom network, despite using algorithms designed to account for cyclic relationships. The CCI algorithm predominantly produced bidirectional arrows, while FCI primarily generated directional arrows, yet neither displayed patterns indicative of definitive cyclic structures. Addressing cyclic relationships is particularly challenging with observational datasets alone. Future research could benefit from refined datasets that include intervention data. Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow* [@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024] are specifically designed to utilize both observational and intervention data to uncover potential cycles.
Additionally, if time-series data becomes available, corresponding methods, as described above, could be applied to capture repetitive patterns in variable interactions that might suggest cyclic structures.

@@

Lastly, regarding conditional independence (CI) testing, the differences between the graphs generated by Gaussian CI and RCoT underscore the methodological sensitivities in detecting causal relationships. Gaussian CI produced denser graphs, which is somewhat counterintuitive, as the Gaussian CI's strict linearity assumption would typically result in fewer detected relationships, not more.
A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data. In contrast, RCoT, free from such assumptions, may better capture these patterns under such conditions. However, while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data. Specifically, the RBF kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions. In such cases, the distances between discrete points may fail to convey meaningful similarity information. As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity.
Future research could address these issues by exploring a broader range of CI testing approaches. One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning]. A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods.
By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches. This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures.

@@

<!-- Implications for Research and Practice -->
Despite its limitations, this study marks a meaningful step toward understanding the mechanisms linking precarity factors and depression. By applying causal discovery methods, it moves beyond traditional association-based analyses, providing insights that can inform more precise and targeted interventions.
While the resulting graphs are preliminary and contain unresolved ambiguities, they offer a valuable starting point for leveraging causal discovery tools to investigate the causal interplay between depression and precarity factors. A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed by **our comp-model paper**. By simulating intervention effects, such models could provide more realistic insights into how targeted actions might influence symptom networks and precarity factors over time. For example, interventions focused on improving sleep hygiene or alleviating guilt could be evaluated for their cascading effects on employment and social relationships, offering actionable guidance for designing population-level mental health strategies.
<!-- Conclusion -->
As one of the early applications of causal discovery tools to the complex dynamics of depression and precarity factors, this study lays a foundation for future research. We hope it inspires further refinement of these methods and ultimately contribute to more effective solutions for alleviating depression and improving societal well-being.

>>>>

# Results
## Depression as sum score
The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precarity factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*) under two different setups: (a) using both Gaussian CI test and RCoT, and (b) using RCoT alone. Black edges represent consistent causal relationships identified by both FCI and CCI, while gray dashed edges denote inconsistent relationships that vary between the two methods. The inconsistent endpoints of the gray edges are marked with circles.

In graph (a), the key pathways suggest that employment precarity (*P.emp*) and social precarity (*P.soc*) do not cause depression (*PHQsum*). *P.soc* does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder. Additionally, *P.emp* and *S.fin* are identified as non-causes of housing precarity (*P.hou*). Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.

Graph (b), derived solely from the nonparametric RCoT method. Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*. 
The relationship between stressors and depression remains unresolved between the two algorithms. However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders. Employment and social precarity may also be influenced by depression, either directly or through latent variables. Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.

@@

The present study examined the causal relationships between precarity factors and depression using causal discovery algorithms, incorporating both sum score and symptom-level analyses. By analyzing PHQ sum scores alongside individual symptom scores, and comparing aggregated precarity scores with individual precarity factors, we gained a comprehensive perspective on how different aspects of precarity influence depression. This dual approach highlighted nuanced pathways that may be overlooked when focusing solely on aggregate measures, while also capturing broader interactions that might remain undetected in analyses limited to individual factors.

@@

<!-- Limitations -->
Several limitations should be acknowledged when interpreting these findings. The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data. These uncertainties highlight the need for further research, particularly with higher-resolution datasets, such as time-series data.  Incorporating time-series data could capture temporal dynamics between precarity factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time. Methods like *PCMCI* [@runge2019detecting], *tsFCI* [@entner2010causal], and other time-series adaptations of causal discovery algorithms could offer a more robust framework for addressing temporal dependencies and refining the analysis.

Another notable limitation is the absence of clear evidence for cycles within the symptom network, despite employing algorithms designed to account for cyclic relationships. The CCI algorithm primarily produced bidirectional arrows, while FCI predominantly generated directional arrows, yet neither demonstrated patterns indicative of definitive cyclic structures. This suggests that addressing cyclic relationships is particularly challenging when relying solely on observational datasets. Future research could benefit from refined datasets that incorporate intervention data. Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow *[@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024], which utilize both observational and intervention data, may uncover potential cycles more effectively.

@@

Lastly, regarding conditional independence (CI) testing, the differences between the graphs generated by Gaussian CI and RCoT underscore the methodological sensitivities in detecting causal relationships. Gaussian CI produced denser graphs, which is somewhat counterintuitive, as the Gaussian CI's strict linearity assumption would typically result in fewer detected relationships, not more.
A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data. In contrast, RCoT, free from such assumptions, might better capture independence patterns under such conditions.
However, RCoT is not without its limitations. While technically non-parametric, its performance can be influenced by the distributional characteristics of the data. The Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle with discrete or mixed datasets, where distances between discrete points may fail to convey meaningful similarity [@howlett2001radial]. This limitation could explain why RCoT might miss certain dependencies, particularly in the HELIUS dataset, where some variables lack smooth continuity.
Future research could tackle these challenges by developing hybrid kernels designed for mixed datasets, integrating both continuous and discrete variables into RCoT-like methods. This would have the potential to improve the reliability of findings, especially when working with datasets characterized by complex and heterogeneous structures.

<!-- However, it could be also that RCoT performed somewhat suboptimal. while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data. Specifically, the Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions. In such cases, the distances between discrete points may fail to convey meaningful similarity information. As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity like in our HELIUS dataset. -->

<!-- Future research could address these issues by exploring a broader range of CI testing approaches. One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning]. A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods. -->
<!-- By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches. This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures. -->

@@

<!-- Implications for Research and Practice -->
Despite its limitations, this study marks a meaningful step toward understanding the mechanisms linking precarity factors and depression. By applying causal discovery methods, it moves beyond traditional association-based analyses, providing insights that can inform more precise and targeted interventions.
While the resulting graphs are preliminary and contain unresolved ambiguities, they offer a valuable starting point for leveraging causal discovery tools to investigate the causal interplay between depression and precarity factors. A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed in **our comp-model paper**. By simulating intervention effects, such models could provide more realistic insights into how targeted actions might influence symptom networks and precarity factors over time. For example, interventions focused on improving sleep hygiene or alleviating guilt could be evaluated for their cascading effects on employment and social relationships, offering actionable guidance for designing population-level mental health strategies.
<!-- Conclusion -->
As one of the early applications of causal discovery tools to the complex dynamics of depression and precarity factors, this study lays a foundation for future research. We hope it inspires further refinement of these methods and ultimately contribute to more effective solutions for alleviating depression and improving societal well-being.

### 2025-01-04 draft_v1/draft_v1.tex
update_discussion_limitations

\date{2025-01-03}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

The sum score graphs provide a high-level summary of how precarity
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precarity factors (\emph{P.hou}, \emph{P.emp},
\emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the depression sum score
(\emph{PHQsum}) under two different setups: (a) using both Gaussian CI
test and RCoT, and (b) using RCoT alone. Black edges represent
consistent causal relationships identified by both FCI and CCI, while
gray edges denote inconsistent relationships that vary between the two
methods. The edge endpoints of inconsistent gray edges are marked with
circles. In graph (a), the key pathways suggest that employment
precarity (\emph{P.emp}) and social precarity (\emph{P.soc}) do not
cause depression (\emph{PHQsum}). Social precarity does not cause recent
relational stress (\emph{S.rel}) or financial stress (\emph{S.fin}), and
these stressors are likely related by a latent confounder. Additionally,
\emph{P.emp} and \emph{S.fin} are identified as non-causes of housing
precarity (\emph{P.hou}). Both FCI and CCI detect a dependency between
\emph{P.emp} and \emph{S.fin}, but they disagree on the direction of the
relationship, leaving it unclear whether \emph{S.fin} causes
\emph{P.emp} or if the connection is mediated by an unmeasured
confounder. A similar ambiguity exists in the relationships between
\emph{S.rel} and \emph{PHQsum} and between \emph{S.fin} and
\emph{PHQsum}, with the methods diverging on whether these stressors
influence depression or are linked through latent variables.

Graph (b), derived solely from the nonparametric RCoT test. Both graphs
consistently identify that employment precarity is not a cause of
housing precarity or depression, and that social precarity does not
cause depression or financial stress. The relationship between stressors
and depression remains unresolved between the two algorithms. However,
RCoT provides greater confidence that depression is not the cause of
financial stress and suggests a stronger likelihood that the financial
stress may contribute to depression or that their relationship is
mediated by a latent confounder. The role of relational stress has
become less pronounced, as the edge between \emph{P.soc} and
\emph{S.rel} is omitted, and the relationship between \emph{S.rel} and
\emph{S.fin} is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of recent
stressors as being closely linked with depression, either as causes or
through latent confounders. Employment and social precarity may also be
influenced by depression, either directly or through latent variables.
Lastly, housing precarity does not directly impact depression but
remains connected to employment precarity and financial stress, either
directly or through a latent confounder.

@@

Several limitations should be acknowledged when interpreting these
findings. The prevalence of bidirectional arrows and circle-marked
endpoints in the graphs reflects unresolved ambiguities in the causal
relationships suggested by the data. These uncertainties underscore the
need for further research, ideally incorporating datasets with a higher
proportion of symptomatic individuals. The HELIUS dataset, being
predominantly composed of asymptomatic samples, posed challenges in
identifying clear causal directions, particularly among symptom nodes.
Future studies could address these limitations by incorporating
time-series data to leverage temporal information about the
relationships between precarity factors and depressive symptoms.
Time-series data could provide time-specific insights and track how
these relationships evolve over time. For instance, methods such as
\emph{PCMCI} (Runge et al., 2019) and \emph{tsFCI} (Entner \& Hoyer,
2010), along with other time-series adaptations of causal discovery
algorithms, could help better account for temporal dependencies and
refine the analysis.

Another limitation is the lack of clear evidence for cycles within the
symptom network, despite using algorithms designed to account for cyclic
relationships. The CCI algorithm predominantly produced bidirectional
arrows, while FCI primarily generated directional arrows, yet neither
displayed patterns indicative of definitive cyclic structures.
Addressing cyclic relationships is particularly challenging with
observational datasets alone. Future research could benefit from refined
datasets that include intervention data. Methods such as \emph{LLC}
(Hyttinen et al., 2012), \emph{NODGAS-Flow} (Sethuraman et al., 2023),
and the recently developed \emph{Bicycle} algorithm (Rohbeck et al.,
2024) are specifically designed to utilize both observational and
intervention data to uncover potential cycles. Additionally, if
time-series data becomes available, corresponding methods, as described
above, could be applied to capture repetitive patterns in variable
interactions that might suggest cyclic structures.

Lastly, regarding conditional independence (CI) testing, the differences
between the graphs generated by Gaussian CI and RCoT underscore the
methodological sensitivities in detecting causal relationships. Gaussian
CI produced denser graphs, which is somewhat counterintuitive, as the
Gaussian CI's strict linearity assumption would typically result in
fewer detected relationships, not more. A possible explanation for this
discrepancy is that Gaussian CI's reliance on partial correlations may
overestimate relationships when specific non-linear dependencies exist
in the data. In contrast, RCoT, free from such assumptions, may better
capture these patterns under such conditions. However, while RCoT is
technically non-parametric, its performance can still be influenced by
the distributional characteristics of the data. Specifically, the RBF
kernel, optimized for continuous data with smooth transitions, may
struggle to capture relationships in datasets with discrete or mixed
distributions. In such cases, the distances between discrete points may
fail to convey meaningful similarity information. As a result, RCoT
might miss certain dependencies, particularly when variables in the
dataset lack smooth continuity. Future research could address these
issues by exploring a broader range of CI testing approaches. One
traditional approach to handle this is discretizing variables and use
\(G^2\) test, which may better capture dependencies in non-continuous
datasets (Dojer, 2016; Neapolitan et al., 2004). A more promising
direction, however, lies in the development of hybrid kernels tailored
for mixed datasets, effectively integrating both continuous and discrete
variables into RCoT-like methods. By systematically employing and
comparing a wider variety of CI testing techniques, researchers could
gain more robust insights and mitigate the limitations inherent in
specific approaches. This broader exploration holds the potential to
enhance the reliability of findings, particularly in datasets with
complex and heterogeneous structures.

Despite its limitations, this study marks a meaningful step toward
understanding the mechanisms linking precarity factors and depression.
By applying causal discovery methods, it moves beyond traditional
association-based analyses, providing insights that can inform more
precise and targeted interventions. While the resulting graphs are
preliminary and contain unresolved ambiguities, they offer a valuable
starting point for leveraging causal discovery tools to investigate the
causal interplay between depression and precarity factors. A promising
next step would involve integrating these causal structures into
computational models, such as the symptom dynamic model proposed by
\textbf{our comp-model paper}. By simulating intervention effects, such
models could provide more realistic insights into how targeted actions
might influence symptom networks and precarity factors over time. For
example, interventions focused on improving sleep hygiene or alleviating
guilt could be evaluated for their cascading effects on employment and
social relationships, offering actionable guidance for designing
population-level mental health strategies. As one of the early
applications of causal discovery tools to the complex dynamics of
depression and precarity factors, this study lays a foundation for
future research. We hope it inspires further refinement of these methods
and ultimately contribute to more effective solutions for alleviating
depression and improving societal well-being.

@@

>>>>

\date{2025-01-04}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

The sum score graphs provide a high-level summary of how precarity
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precarity factors (\emph{P.hou}, \emph{P.emp},
\emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the depression sum score
(\emph{PHQsum}) under two different setups: (a) using both Gaussian CI
test and RCoT, and (b) using RCoT alone. Black edges represent
consistent causal relationships identified by both FCI and CCI, while
gray dashed edges denote inconsistent relationships that vary between
the two methods. The inconsistent endpoints of the gray edges are marked
with circles.

In graph (a), the key pathways suggest that employment precarity
(\emph{P.emp}) and social precarity (\emph{P.soc}) do not cause
depression (\emph{PHQsum}). \emph{P.soc} does not cause recent
relational stress (\emph{S.rel}) or financial stress (\emph{S.fin}), and
these stressors are likely related by a latent confounder. Additionally,
\emph{P.emp} and \emph{S.fin} are identified as non-causes of housing
precarity (\emph{P.hou}). Both FCI and CCI detect a dependency between
\emph{P.emp} and \emph{S.fin}, but they disagree on the direction of the
relationship, leaving it unclear whether \emph{S.fin} causes
\emph{P.emp} or whether a latent confounder mediates their relationship.
Similar ambiguities are found in the connections between \emph{S.rel}
and \emph{PHQsum} and between \emph{S.fin} and \emph{PHQsum}.

Graph (b), derived solely from the nonparametric RCoT method. Both
graphs consistently identify that \emph{P.emp} is not a cause of
\emph{P.hou} or \emph{PHQsum}, and that \emph{P.soc} does not cause
\emph{PHQsum} or \emph{S.fin}. The relationship between stressors and
depression remains unresolved between the two algorithms. However, the
graph derived from RCoT provides greater confidence that depression is
not the cause of \emph{S.fin} and suggests a stronger likelihood that
the financial stress may contribute to depression or that their
relationship is mediated by a latent confounder. The role of relational
stress has become less pronounced, as the edge between \emph{P.soc} and
\emph{S.rel} is omitted, and the relationship between \emph{S.rel} and
\emph{S.fin} is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of stressors
like \emph{S.fin} and \emph{S.rel}, which are closely tied to
depression, either causally or via latent confounders. Employment and
social precarity may also be influenced by depression, either directly
or through latent variables. Lastly, housing precarity does not directly
impact depression but remains connected to employment precarity and
financial stress, likely mediated by direct or latent pathways.

@@

Several limitations should be acknowledged when interpreting these
findings. The prevalence of bidirectional arrows and circle-marked
endpoints in the graphs reflects unresolved ambiguities in the causal
relationships suggested by the data. These uncertainties highlight the
need for further research, particularly with higher-resolution datasets,
such as time-series data. Incorporating time-series data could capture
temporal dynamics between precarity factors and depressive symptoms,
providing time-specific insights and revealing how these relationships
evolve over time. Methods like \emph{PCMCI} (Runge et al., 2019),
\emph{tsFCI} (Entner \& Hoyer, 2010), and other time-series adaptations
of causal discovery algorithms could offer a more robust framework for
addressing temporal dependencies and refining the analysis.

Another notable limitation is the absence of clear evidence for cycles
within the symptom network, despite employing algorithms designed to
account for cyclic relationships. The CCI algorithm primarily produced
bidirectional arrows, while FCI predominantly generated directional
arrows, yet neither demonstrated patterns indicative of definitive
cyclic structures. This suggests that addressing cyclic relationships is
particularly challenging when relying solely on observational datasets.
Future research could benefit from refined datasets that incorporate
intervention data. Methods such as \emph{LLC} (Hyttinen et al., 2012),
\emph{NODGAS-Flow }(Sethuraman et al., 2023), and the recently developed
\emph{Bicycle} algorithm (Rohbeck et al., 2024), which utilize both
observational and intervention data, may uncover potential cycles more
effectively.

Lastly, regarding conditional independence (CI) testing, the differences
between the graphs generated by Gaussian CI and RCoT underscore the
methodological sensitivities in detecting causal relationships. Gaussian
CI produced denser graphs, which is somewhat counterintuitive, as the
Gaussian CI's strict linearity assumption would typically result in
fewer detected relationships, not more. A possible explanation for this
discrepancy is that Gaussian CI's reliance on partial correlations may
overestimate relationships when specific non-linear dependencies exist
in the data. In contrast, RCoT, free from such assumptions, might better
capture independence patterns under such conditions. However, RCoT is
not without its limitations. While technically non-parametric, its
performance can be influenced by the distributional characteristics of
the data. The Gaussian RBF (radial basis function) kernel, optimized for
continuous data with smooth transitions, may struggle with discrete or
mixed datasets, where distances between discrete points may fail to
convey meaningful similarity (Howlett, 2001). This limitation could
explain why RCoT might miss certain dependencies, particularly in the
HELIUS dataset, where some variables lack smooth continuity. Future
research could tackle these challenges by developing hybrid kernels
designed for mixed datasets, integrating both continuous and discrete
variables into RCoT-like methods. This would have the potential to
improve the reliability of findings, especially when working with
datasets characterized by complex and heterogeneous structures.

Despite its limitations, this study marks a meaningful step toward
understanding the mechanisms linking precarity factors and depression.
By applying causal discovery methods, it moves beyond traditional
association-based analyses, providing insights that can inform more
precise and targeted interventions. While the resulting graphs are
preliminary and contain unresolved ambiguities, they offer a valuable
starting point for leveraging causal discovery tools to investigate the
causal interplay between depression and precarity factors. A promising
next step would involve integrating these causal structures into
computational models, such as the symptom dynamic model proposed in
\textbf{our comp-model paper}. By simulating intervention effects, such
models could provide more realistic insights into how targeted actions
might influence symptom networks and precarity factors over time. For
example, interventions focused on improving sleep hygiene or alleviating
guilt could be evaluated for their cascading effects on employment and
social relationships, offering actionable guidance for designing
population-level mental health strategies. As one of the early
applications of causal discovery tools to the complex dynamics of
depression and precarity factors, this study lays a foundation for
future research. We hope it inspires further refinement of these methods
and ultimately contribute to more effective solutions for alleviating
depression and improving societal well-being.

@@

\bibitem[\citeproctext]{ref-howlett2001radial}
Howlett, R. J. (2001). \emph{Radial basis function networks 1: Recent
developments in theory and applications}.
